The Concurrent Calculi Formalisation Benchmark

### 2024-02-19 short-paper/short-paper.tex
Reorganise the short paper.
This undoes the idea of splitting the related work in two sections,
since this never really worked out well.
This also cleans up the introduction by moving things elsewhere.

The sections about each challenge are not integrated yet, and neither
are the sections on the overall design and possible extensions.

\todo{These paragraphs are still quite messy and unfocused. They should focus on the motivation, i.e.\ mention what others have perceived as problems.}

\alb{I really do not like the duplication of the discussion of the
  three issues here and in Sec 2. I understand the desire to highlight
  noted issues, but it does not work. Some citations are unconvincing. Mostly, we are anticipating the description of the challenges. Let's unify this}
\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.

\todo{Insert related work focusing on noted issues with mechanising linearity.}

\paragraph{Scope extrusion.}
The encoding of binding constructs has been extensively studied
w.r.t.\ the $\lambda$-calculi, but process calculi present additional
challenges.  $\pi$-calculi typically include several different binding
constructs: inputs bind a received name or value, recursive processes
bind recursion variables, and restrictions bind names.  The first two
act similarly to the binders known from $\lambda$-calculi, but
restrictions may be more challenging due to scope extrusion.

\todo{Insert related work focusing on noted issues when mechanising scope extrusion.}

\paragraph{Coinduction.} \todo{explain more what the challenge does.}
\lore{I have tried this:
  the first paragraph motivates (``why is this difficult?''), while
  the second talks about the challenge.}
Coinduction is fundamental to establish properties of processes that
may never terminate, which is a typical situation.
Common uses for coinduction are to define strong and weak versions of
bisimulation, barbed notions of bisimulation, and prove trace equivalence.
On the other hand, coinduction in proof assistants
is notoriously difficult and its representation
is non-uniform across developments.
E.g., \cite{Castro-Perez2021} notes that the most challenging
part of their mechanisation of multiparty session types
was the convoluted definition of infinite behaviours;
\cite{Pohjola2022} points out that the main difficulty
of the proofs concerning bisimulation lies
in dealing with existential quantifiers; and
\cite{Bengtson2016} wonders whether parametrised
coinduction as developed in Coq \cite{Hur2013} would
have helped their Isabelle mechanisation.

We base our challenge around bisimulation: a fundamental coinductively
established property for equating the
observed behaviour of potentially infinite processes. 
In this challenge, we study the notion of strong barbed bisimulation
(and bisimilarity, as in \cite{picalcbook})
over a fragment of the $\pi$-calculus and,
in particular, we focus on obtaining,
from strong barbed bisimilarity,
a congruence with respect to
parallel composition and substitution.
%
%% The introduction of bisimulation
%% is one of the first steps, when studying the
%% (observable) behaviour of process calculi, both
%% in classical textbooks (e.g., \cite{picalcbook})
%% and in existing 
%% mechanisations
%% \cite{Thiemann2019,Honsell2001,Bengtson2016,Kahsai2008,Gay2020,
%% Hirschkoff1997,Bengtson2009,Maksimovic2015}.

%% , where we see a variety of
%% different formalisation choices.
%% As a significant example, we observe that,
%% while many authors rely
%% on the native support for coinduction offered by
%% proof assistants
%% \cite{Thiemann2019,Honsell2001,Bengtson2016,Kahsai2008,Gay2020},
%% many others prefer a more ``set-theoretic''
%% approach
%% \cite{Hirschkoff1997,Bengtson2009,Maksimovic2015}.

%Common uses for coinduction are to define strong and weak versions of
%bisimulation, barbed notions of bisimulation, prove trace equivalence,
%etc.

%In this challenge we study the notion of strong barbed bisimulation
%over a fragment of the $\pi$-calculus.

\todo{Insert related work focusing on noted issues when mechanising coinduction.}

\alb{This subsection does not make a lot of sense here -- introductions do not have subsections. Let's move it in ``Design'' }
\subsection{Goals}% and how to contribute}
Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
\alb{Should we stress we are interested in the \textbf{meta-theory} of process calculi, not}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

% - Design of the benchmark
%   - We first list some general design remarks, then for each challenge
%   - General remarks
%     - We do not aim to be comprehensive (list other topics)
%     - However, we believe the best practices developed could also be of use in those topics
%     - We base our challenges on basic textbook theory with interesting proof techniques
%     - The challenges are independent (maybe note ``failure'' of POPLMark that everyone only did the first part)
%     - The challenges are as minimal as possible
%   - Linearity
%     - Type preservation for session types
%     - The main issue is context splitting
%     - Session types are pervasive in the field and simple to understand
%     - Binding two names makes duality obvious
%     - Brief overview of approaches and issues with them
%   - Scope extrusion
%     - Process calculi have several binders, but only restriction is different from ``normal'' ones
%     - The main issue is scope extrusion
%     - The pros and cons of established techniques for handling binders are different for process calculi, and it is unclear how exactly
%     - Brief overview of approaches and issues with them
%   - Coinduction
%     - Strong barbed bisimulation and congruence
%     - The main issue is the encoding of coinductive up-to techniques
%     - Strong equivalences to simplify (we don't have to abstract over internal transitions)
%     - Why barbed equivalences?
%     - No delegation to simplify
%     - Brief overview of approaches and issues with them
%   - Evaluation criteria
%     - We care not only about solutions, but about experiences and comparisons
%     - Mechanisation overhead
%     - Adequacy of the formal statements
%     - Cost of entry for tools and techniques
%     - Slight changes are OK if they are justified and adequate
\section{Design of the Benchmark}\label{sec:design-discussion}
\todo{Add references to support our claims and decisions in this whole section. Here is where we should mention existing approaches.}

In this section, we outline the factors considered in designing the benchmark challenges.
We begin with some general remarks, then describe the individual design considerations for each challenge problem, and the evaluation criteria for solutions.

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is
not meant to be comprehensive: applications such as multiparty session
types~\cite{10.1145/2827695,10.1145/3290343},
choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation
types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696},
or encodings between different
calculi~\cite{DBLP:journals/iandc/Gorla10,%,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,
  DBLP:conf/ecoop/ScalasDHY17}
  %,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407}
are not directly covered.

\alb{I've commented some citations about
  encodings. Is it a good idea to start the section describing what we \textbf{do not do}?}

Still, these (and other) applications need the basic
techniques that our challenge problems emphasize:
Aligned with design goal~\ref{item:goal-reusability}, our problems are rooted in the fundamental meta-theory of process calculi, without necessitating the creation of novel theory. These problems center around well-established theorems and results, showcasing interesting proof techniques that can be leveraged in future endeavors.
% as per design goal
% \ref{item:goal-reusability}, our problems are drawn from the basic
% meta-theory of process calculi (without requiring the development of
% new theory), and focus on well-known theorems and results that involve
% interesting proof techniques that may be reused in further work. 
To
achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem is %should be
reasonably small and easily understandable with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasize, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible benefits and insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.

\subsubsection{Linearity.}

\alb{I would put the challenge before the literature}
For our challenge on linear reasoning we have chosen a type safety theorem
for a system using session types, combining as usual type preservation with absence of errors.
% , whose use are common in concurrency theory.
This allows us to introduce linear reasoning with few definitions.
With our objective focused on linearity, we forego channel delegation
and opt for a reduction approach rather than employing a transition
system semantics.  Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names, highlighting
their duality within the type system.  We have chosen a simple notion
of well-formedness, and do not  consider
deadlocks. While more sophisticated notions of well-formedness are
interesting, the proofs of these properties would distract from the
core linear aspects of the challenge.

@@

\todo{Insert discussion/related work focusing on existing techniques and how our benchmark design makes it possible to compare them.}

\alb{This needs some work: we do not describe the challenge at all and the citations are not the one I had in mind}
\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders in restrictions and input.

@@

\todo{Insert discussion/related work focusing on existing techniques and how our benchmark design makes it possible to compare them.}

@@

\subsubsection{Coinduction.}
\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}

@@

\lore{Should we mention that Isabelle offers
  the command \texttt{coinduction\_upto}? Is it related?}
\todo{Insert discussion/related work focusing on existing techniques and how our benchmark design makes it possible to compare them.}
\lore{I have added something below... Any advice on what more we need?}

%As a significant example, we observe that,
While many authors rely on the native support for coinduction offered
by proof assistants
\cite{Honsell2001,Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, many
others prefer a more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}.
\alb{Actually, Honsell gives up with guarded coinduction and switches to Park's half way}

%% \lore{Need to integrate the following with the above.}
%% Different process-calculi
%% mechanisations
%% implement different techniques;
%% most of the time these 
%% are ad-hoc efforts to achieve a specific result.
%% \lore{cite and substantiate.}
%% In parallel, although most proof assistants
%% offer support for coinduction in their standard libraries,
%% research has mainly focused on the foundations
%% of the specific system, more than on
%% applicability.
%% \lore{cite and substantiate.}

\alb{This should not be a subsubsection at the same level of coinduction etc}
\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering
questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore
interested not only in the solutions, but also in the experience of
solving the challenges with the chosen approach.  Solutions to our
challenges should be compared on three axes:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.

@@

On the mechanisation side, the present challenge proposes mechanising the proofs of theorems.
An interesting avenue to explore would be to take advantage of other proof assistant features like extracting certified implementations.
Additionally, while our current challenge is agnostic about proof automation, a more specific challenge could be designed with the objective of automating aspects of the proofs.
Finally, a challenge could propose the integration with other formal reasoning tools, such as model checkers.
These tools are in common use in the field; combining automated proofs with proof assistants offers the potential to ease the path towards larger proofs.

>>>>

\section{Design of the Benchmark}\label{sec:design-discussion}
In this section, we outline the factors considered in designing the benchmark challenges.
We begin with some general remarks, then describe the individual design considerations for each challenge problem, and the evaluation criteria for solutions.

Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
\alb{Should we stress we are interested in the \textbf{meta-theory} of process calculi, not}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\todo{Refer to the future benchmark section about other things one could do, but they still all need the basics}
Still, these (and other) applications need the basic
techniques that our challenge problems emphasize:
Aligned with design goal~\ref{item:goal-reusability}, our problems are rooted in the fundamental meta-theory of process calculi, without necessitating the creation of novel theory. These problems center around well-established theorems and results, showcasing interesting proof techniques that can be leveraged in future endeavors.
To achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem is
reasonably small and easily understandable with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasize, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible benefits and insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.

\subsubsection{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.

\alb{We need more details about what the challenge is}
For our challenge on linear reasoning we have chosen a type safety theorem
for a system using session types, combining as usual type preservation with absence of errors.
This allows us to introduce linear reasoning with few definitions.
With our objective focused on linearity, we forego channel delegation
and opt for a reduction approach rather than employing a transition
system semantics.  Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names, highlighting
their duality within the type system.  We have chosen a simple notion
of well-formedness, and do not  consider
deadlocks. While more sophisticated notions of well-formedness are
interesting, the proofs of these properties would distract from the
core linear aspects of the challenge.

@@

\alb{This needs some work: we do not describe the challenge at all and the citations are not the one I had in mind}
\subsubsection{Scope extrusion.}
The encoding of binding constructs has been extensively studied
w.r.t.\ the $\lambda$-calculi, but process calculi present additional
challenges.  $\pi$-calculi typically include several different binding
constructs: inputs bind a received name or value, recursive processes
bind recursion variables, and restrictions bind names.  The first two
act similarly to the binders known from $\lambda$-calculi, but
restrictions may be more challenging due to scope extrusion.

\alb{We need more details about what the challenge is}
This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders in restrictions and input.

@@

@@

\subsubsection{Coinduction.}
Coinduction is fundamental to establish properties of processes that
may never terminate, which is a typical situation.
Common uses for coinduction are to define strong and weak versions of
bisimulation, barbed notions of bisimulation, and prove trace equivalence.
On the other hand, coinduction in proof assistants
is notoriously difficult and its representation
is non-uniform across developments.
E.g., \cite{Castro-Perez2021} notes that the most challenging
part of their mechanisation of multiparty session types
was the convoluted definition of infinite behaviours;
\cite{Pohjola2022} points out that the main difficulty
of the proofs concerning bisimulation lies
in dealing with existential quantifiers; and
\cite{Bengtson2016} wonders whether parametrised
coinduction as developed in Coq \cite{Hur2013} would
have helped their Isabelle mechanisation.

We base our challenge around bisimulation: a fundamental coinductively
established property for equating the
observed behaviour of potentially infinite processes. 
In this challenge, we study the notion of strong barbed bisimulation
(and bisimilarity, as in \cite{picalcbook})
over a fragment of the $\pi$-calculus and,
in particular, we focus on obtaining,
from strong barbed bisimilarity,
a congruence with respect to
parallel composition and substitution.

\alb{We need more details about what the challenge is}
\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}

@@

\lore{Should we mention that Isabelle offers
  the command \texttt{coinduction\_upto}? Is it related?}
\todo{Explain the the point of the congruence theorem is that it should be ``easy'' once you have the other one}

While many authors rely on the native support for coinduction offered
by proof assistants
\cite{Honsell2001,Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, many
others prefer a more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}.
\alb{Actually, Honsell gives up with guarded coinduction and switches to Park's half way}

\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering
questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore
interested not only in the solutions, but also in the experience of
solving the challenges with the chosen approach.  Solutions to our
challenges should be compared on three axes:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.

@@

Like the POPLMark and POPLMark Reloaded challenges, our challenge is
not meant to be comprehensive: applications such as multiparty session
types~\cite{10.1145/2827695,10.1145/3290343},
choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation
types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696},
or encodings between different
calculi~\cite{DBLP:journals/iandc/Gorla10, DBLP:conf/ecoop/ScalasDHY17} are not directly covered.\todo{Should we cite other/different encoding papers? Which ones are the ``standard'' citations?}

\todo{This is a list of interesting research problems, but they are not challenges per se. Trim down.}
On the mechanisation side, the present challenge proposes mechanising the proofs of theorems.
An interesting avenue to explore would be to take advantage of other proof assistant features like extracting certified implementations.
Additionally, while our current challenge is agnostic about proof automation, a more specific challenge could be designed with the objective of automating aspects of the proofs.
